Cooling Requirements for the Vertical Shear Instability in Protoplanetary Disks

### 2015-04-23 appendix.tex

In the vertically global shearing box (VGSB) approximation, we set
$\hat{g}_c=0$ in Eq. \ref{lin_mass} --- \ref{lin_energy}. We then eliminate
the horizonatal velocity perturbations ($\delta v_x,\, \delta v_y$) to
obtain  
%we can eliminate variables 
%to obtain a pair or ordinary differential
%equations for $W,Q$ as 
\begin{align}
  & \ii\upsilon\delta v_z = \frac{dW}{dz} + \frac{d\ln{\rho}}{d z}\left(W-Q\right),\label{ode_vz} \\
  &\frac{\upsilon^2}{c_s^2}Q + \frac{\upsilon^2k_x^2}{D}W =\ii\upsilon \left(\frac{\ii
      k_x r}{D}\frac{d\Omega^2}{dz} -
    \frac{d\ln{\rho}}{dz}\right)\delta v_z - \ii\upsilon \frac{d\delta v_z}{dz},
  % \left[\frac{dW}{dz} + \frac{d\ln{\rho}}{dz}\left(W-Q\right)\right] 
  % \notag\\
  % &- \frac{d^2W}{dz^2} - \frac{d\ln{\rho}}{dz}\left(\frac{dW}{dz} -
  %   \frac{dQ}{dz}\right) - \frac{d^2\ln{\rho}}{dz^2}\left(W-Q\right),
  \label{ode_w}\\
%  &\sigma^2W - \frac{\gamma}{\Gamma}\sigma^2 Q +
%  \frac{\ii\sigma}{t_c}\left(W-\frac{Q}{\Gamma}\right) 
%  =\ii \sigma c_s^2\frac{d\ln{\rho}}{dz}\left(\frac{\gamma}{\Gamma} -
%    1\right)\delta v_z,  
  &\upsilon^2W - \gamma\upsilon^2 Q +
  \frac{\ii\upsilon}{t_c}\left(W-Q\right) 
  =\ii \upsilon c_s^2\frac{d\ln{\rho}}{dz}\left(\gamma -
    1\right)\delta v_z,
  % 
  % \left[\frac{dW}{dz} + \frac{d\ln{\rho}}{dz}\left(W-Q\right)\right],
  \label{ode_Q} 
\end{align}
(note that $\Gamma=1$ has been used) where
\begin{align}
  D \equiv \kappa^2 - \upsilon^2,
\end{align} 
and we have replaced the notation $\p_z$ by $d/dz$ since we are
considering solutions at a fixed radius.

@@

The background radial disk structure explicitly appears 
in the linearized mass, radial momentum and energy equations
(Eq. \ref{lin_mass}, \ref{lin_vx} and \ref{lin_energy}) as the terms
proportional to $\hat{g}_c$. Nominally $\hat{g}_c=1$.      
In the VGSB (and SSB) these terms are neglected by setting
$\hat{g}_c=0$. For a power-law disk, these radial gradients  
are $O(r^{-1})$, and they appear in comparison with terms of
$O(k_x)$. The neglected terms therefore have a relative magnitude of
$O(h/\khat)$, which is small for thin disks ($h\ll1$)
and/or small radial wavelengths ($\khat\gg 1$).

@@

%However, background radial gradient terms 
%are ignored elsewhere in the VGSB (which is done by setting
%$\hat{g}_c=0$).

% \begin{align}
%   \ii \sigma W  &=c _s^2(r_0, z)\left( \left.\frac{\p\ln\rho}{\p
%       z}\right|_0 \delta v_z + \frac{\gamma}{\Gamma} \frac{d\delta
%     v_z}{dz}\right) + \left[\ii k_xc_s^2(r_0,z)
%   \frac{\gamma}{\Gamma} + \hat{g}_c\left.\frac{1}{\rho}\frac{\p P}{\p
%       r}\right|_0\right]\delta v_x  +
% \frac{1}{t_c}\left(W-\frac{Q}{\Gamma}\right),\label{gcorr_terms1}\\
% \ii \sigma Q &= c_s^2(r_0,z)\left(\ii k_x + \hat{g}_c
%   \left.\frac{\p\ln{\rho}}{\p r}\right|_0\right)\delta v_x + c _s^2(r_0, z)\left( \left.\frac{\p\ln\rho}{\p
%       z}\right|_0 \delta v_z + \frac{d\delta
%     v_z}{dz}\right),\label{gcorr_terms2} \\
% \ii \sigma \delta v_x & = \ii k_x W  -
% \hat{g}_c\frac{1}{c_s^2(r_0,z)}\left.\frac{1}{\rho}\frac{\p P}{\p
%   r}\right|_0Q - 2\Omega(r_0,z)\delta v_y,\label{gcorr_terms3}
% \end{align}
% where subscript $0$ denotes evaluation at $r=r_0$. We recall terms
% proportional to $\hat{g}_c$ are associated with the background density
% and pressure radial structures, and nominally $\hat{g}_c=1$.

\subsection{Spurious growth of adiabatic perturbations in the VGSB}\label{analytic_adia} 
A limitation of the VGSB is that it cannot be used for 
adiabatic flow, even if the above conditions are satisfied. We explain this by considering vertically isothermal
disks and setting  $\beta\to\infty$ and hence $\chi = 1/\gamma$ in 
Eq. \ref{vertiso_gov_nondim} % (which itself is obtained by setting
% $\hat{g}_c=0$ in Eq. \ref{lin_mass}---\ref{lin_energy})
to give 
\begin{align}
  0 =\dd v_z^{\prime\prime} + \left(1 + \ii h q
    \hat{k}\right)\left(\ln\rho^{\prime}\delta v_z\right)^\prime
  +\left\{\hat{\upsilon}^2\left(\frac{1}{\gamma}+\hat{k}^2\right) 
    -\left(\frac{\gamma-1}{\gamma}\right)\left[\ln\rho^{\prime\prime}+\hat{k}^2\left(1-\frac{\ii h  
          q}{\hat{k}}\right)\ln\rho^{\prime 2}\right]\right\}\delta v_z.\label{adia_iso3}
\end{align}
We multiply Eq. \ref{adia_iso3} by $\rho\delta v_z^*$ and
integrate vertically, assume boundary terms vanish when integrating by
parts, to obtain
\begin{align}
  \hat{\upsilon}^2\left(\frac{1}{\gamma} +
    \hat{k}^2\right)\int_{\zhat_1}^{\zhat_2}\rho|\delta
  v_z|^2 d\zhat 
  =&  \left(\frac{\gamma-1}{\gamma}\right)
  \int_{\zhat_1}^{\zhat_2}\rho|\delta v_z^\prime|^2 d\zhat
  +\frac{1}{\gamma}\int_{\zhat_1}^{\zhat_2}\frac{1}{\rho}|(\rho\delta
  v_z)^\prime|^2 d\zhat\notag\\
&+
  \left(\frac{\gamma-1}{\gamma}\right)\hat{k}^2\left(1-\frac{\ii h
      q}{\hat{k}}\right) \int_{\zhat_1}^{\zhat_2}\rho\ln\rho^{\prime
    2}|\delta v_z|^2 d\zhat
+ \ii h q \hat{k}
  \int_{\zhat_1}^{\zhat_2}\ln\rho^\prime(\rho\delta v_z^*)^\prime
  \delta v_z d\zhat.\label{adia_integral}
\end{align}
% When $q\equiv0$, all the terms on the right-hand-side (RHS) are real. Then
% $\hat{\upsilon}^2$ is real, and  $\hat{\upsilon}^2>0$ if $\gamma>1$. As
% expected, a sub-adiabatically stratified disk is stable in the absence
% of vertical shear. 
In the presence of vertical shear $q\neq0$, Eq. \ref{adia_integral}
suggests $\hat{\upsilon}^2$ is complex for real $\khat$, which implies the
possibility of instability, regardless of the value of
$|\gamma-1|$. This is contradicts the second Solberg-Hoiland criterion 
(Eq. \ref{solberg2}), which imply instability for adiabatic
perturbations in a thin, vertically isothermal disk would require 
$|\gamma-1|\ll1$. 
%also need real freq^2 in adia limit 
This is because the VGSB retains the global radial
disk structure responsible for vertical shear
($\p_z\Omega$), but ignores it elsewhere in the linear problem. 
Nevertheless, we demonstrate below
that this inconsistency is unimportant for the VSI, which occurs 
for $\beta\ll1$.   %and small scale

\subsection{Effect of global radial gradients}
In  Fig. \ref{gcorr_compare}, we plot the effect of the global radial gradient terms
proportional to $\hat{g}_c$ in Eq. \ref{lin_mass}---\ref{lin_energy}
by calculating the
fundamental VSI growth rates using three approaches. We compute growth rates from the dispersion
relation Eq. \ref{relax_disp}, which assumes $\hat{g}_c=0$;  from
Eq. \ref{lin_mass}---\ref{lin_energy} with $\hat{g}_c=0$; and from
Eq. \ref{lin_mass}---\ref{lin_energy}  with $\hat{g}_c=1$.

@@

\begin{figure}
  \includegraphics[width=\linewidth,clip=true,trim=0cm 0.0cm 0cm
  0cm]{figures/gcorr_compare} 
  \caption{Growth rate of the fundamental VSI mode as a function of
    the thermal relaxation time $\beta$. The `Analytical (VGSB)' curve
    is calculated from the dispersion relation Eq. \ref{relax_disp};
    the `Numerical (VGSB)' curve is obtained by numerically solving
    the linear problem, Eq. \ref{lin_mass}---\ref{lin_energy}, with $\hat{g}_c=0$
    (which defines the linear VGSB equations); and the `Numerical
    (Full)' curve is obtained by numerically solving 
    Eq. \ref{lin_mass}---\ref{lin_energy}  with $\hat{g}_c=1$, which
    accounts for the radial disk structure.  
    The disk parameters are $(\gamma, \Gamma)=(1.4, 1.011)$ and
    $(p,q,h)=(-1.5,-1,0.05)$. The perturbation
    wavenumber is $\khat=30$. 
    \label{gcorr_compare}}  
\end{figure}

>>>>

In the vertically global shearing box (VGSB) approximation, we set
$\zeta=0$ in Eq. \ref{lin_mass} --- \ref{lin_energy}. We then eliminate
the horizonatal velocity perturbations ($\delta v_x,\, \delta v_y$) to
obtain  
%we can eliminate variables 
%to obtain a pair or ordinary differential
%equations for $W,Q$ as 
\begin{align}
  & \ii\upsilon\delta v_z = \frac{dW}{dz} + \frac{d\ln{\rho}}{d z}\left(W-Q\right),\label{ode_vz} \\
  &\frac{\upsilon^2}{c_s^2}Q + \frac{\upsilon^2k_x^2}{D}W =\ii\upsilon \left(\frac{\ii
      k_x r}{D}\frac{d\Omega^2}{dz} -
    \frac{d\ln{\rho}}{dz}\right)\delta v_z - \ii\upsilon \frac{d\delta v_z}{dz},
  % \left[\frac{dW}{dz} + \frac{d\ln{\rho}}{dz}\left(W-Q\right)\right] 
  % \notag\\
  % &- \frac{d^2W}{dz^2} - \frac{d\ln{\rho}}{dz}\left(\frac{dW}{dz} -
  %   \frac{dQ}{dz}\right) - \frac{d^2\ln{\rho}}{dz^2}\left(W-Q\right),
  \label{ode_w}\\
%  &\sigma^2W - \frac{\gamma}{\Gamma}\sigma^2 Q +
%  \frac{\ii\sigma}{t_c}\left(W-\frac{Q}{\Gamma}\right) 
%  =\ii \sigma c_s^2\frac{d\ln{\rho}}{dz}\left(\frac{\gamma}{\Gamma} -
%    1\right)\delta v_z,  
  &\upsilon^2W - \gamma\upsilon^2 Q +
  \frac{\ii\upsilon}{t_c}\left(W-Q\right) 
  =\ii \upsilon c_s^2\frac{d\ln{\rho}}{dz}\left(\gamma -
    1\right)\delta v_z,
  % 
  % \left[\frac{dW}{dz} + \frac{d\ln{\rho}}{dz}\left(W-Q\right)\right],
  \label{ode_Q} 
\end{align}
(note that $\Gamma=1$ has been used) where
\begin{align}
  D \equiv \kappa^2 - \upsilon^2,
\end{align} 
and we have replaced the notation $\p_z$ by $d/dz$ since we are
considering solutions at a fixed radius.

@@

The background radial disk structure explicitly appears 
in the linearized mass, radial momentum and energy equations
(Eq. \ref{lin_mass}, \ref{lin_vx} and \ref{lin_energy}) as the terms
proportional to $\zeta$. Nominally $\zeta=1$.      
In the VGSB (and SSB) these terms are neglected by setting
$\zeta=0$. For a power-law disk, these radial gradients  
are $O(r^{-1})$, and they appear in comparison with terms of
$O(k_x)$. The neglected terms therefore have a relative magnitude of
$O(h/\khat)$, which is small for thin disks ($h\ll1$)
and/or small radial wavelengths ($\khat\gg 1$).

@@

%However, background radial gradient terms 
%are ignored elsewhere in the VGSB (which is done by setting
%$\zeta=0$).

% \begin{align}
%   \ii \sigma W  &=c _s^2(r_0, z)\left( \left.\frac{\p\ln\rho}{\p
%       z}\right|_0 \delta v_z + \frac{\gamma}{\Gamma} \frac{d\delta
%     v_z}{dz}\right) + \left[\ii k_xc_s^2(r_0,z)
%   \frac{\gamma}{\Gamma} + \zeta\left.\frac{1}{\rho}\frac{\p P}{\p
%       r}\right|_0\right]\delta v_x  +
% \frac{1}{t_c}\left(W-\frac{Q}{\Gamma}\right),\label{gcorr_terms1}\\
% \ii \sigma Q &= c_s^2(r_0,z)\left(\ii k_x + \zeta
%   \left.\frac{\p\ln{\rho}}{\p r}\right|_0\right)\delta v_x + c _s^2(r_0, z)\left( \left.\frac{\p\ln\rho}{\p
%       z}\right|_0 \delta v_z + \frac{d\delta
%     v_z}{dz}\right),\label{gcorr_terms2} \\
% \ii \sigma \delta v_x & = \ii k_x W  -
% \zeta\frac{1}{c_s^2(r_0,z)}\left.\frac{1}{\rho}\frac{\p P}{\p
%   r}\right|_0Q - 2\Omega(r_0,z)\delta v_y,\label{gcorr_terms3}
% \end{align}
% where subscript $0$ denotes evaluation at $r=r_0$. We recall terms
% proportional to $\zeta$ are associated with the background density
% and pressure radial structures, and nominally $\zeta=1$.

\subsection{Spurious growth of adiabatic perturbations in the VGSB}\label{analytic_adia} 
A limitation of the VGSB is that it cannot be used for 
adiabatic flow, even if the above conditions are satisfied. We explain this by considering vertically isothermal
disks and setting  $\beta\to\infty$ and hence $\chi = 1/\gamma$ in 
Eq. \ref{vertiso_gov_nondim} % (which itself is obtained by setting
% $\zeta=0$ in Eq. \ref{lin_mass}---\ref{lin_energy})
to give 
\begin{align}
  0 =\dd v_z^{\prime\prime} + \left(1 + \ii h q
    \hat{k}\right)\left(\ln\rho^{\prime}\delta v_z\right)^\prime
  +\left\{\hat{\upsilon}^2\left(\frac{1}{\gamma}+\hat{k}^2\right) 
    -\left(\frac{\gamma-1}{\gamma}\right)\left[\ln\rho^{\prime\prime}+\hat{k}^2\left(1-\frac{\ii h  
          q}{\hat{k}}\right)\ln\rho^{\prime 2}\right]\right\}\delta v_z.\label{adia_iso3}
\end{align}
We multiply Eq. \ref{adia_iso3} by $\rho\delta v_z^*$ and
integrate vertically, assume boundary terms vanish when integrating by
parts, to obtain
\begin{align}
  \hat{\upsilon}^2\left(\frac{1}{\gamma} +
    \hat{k}^2\right)\int_{\zhat_1}^{\zhat_2}\rho|\delta
  v_z|^2 d\zhat 
  =&  \left(\frac{\gamma-1}{\gamma}\right)
  \int_{\zhat_1}^{\zhat_2}\rho|\delta v_z^\prime|^2 d\zhat
  +\frac{1}{\gamma}\int_{\zhat_1}^{\zhat_2}\frac{1}{\rho}|(\rho\delta
  v_z)^\prime|^2 d\zhat\notag\\
&+
  \left(\frac{\gamma-1}{\gamma}\right)\hat{k}^2\left(1-\frac{\ii h
      q}{\hat{k}}\right) \int_{\zhat_1}^{\zhat_2}\rho\ln\rho^{\prime
    2}|\delta v_z|^2 d\zhat
+ \ii h q \hat{k}
  \int_{\zhat_1}^{\zhat_2}\ln\rho^\prime(\rho\delta v_z^*)^\prime
  \delta v_z d\zhat.\label{adia_integral}
\end{align}
% When $q\equiv0$, all the terms on the right-hand-side (RHS) are real. Then
% $\hat{\upsilon}^2$ is real, and  $\hat{\upsilon}^2>0$ if $\gamma>1$. As
% expected, a sub-adiabatically stratified disk is stable in the absence
% of vertical shear. 
In the presence of vertical shear $q\neq0$, Eq. \ref{adia_integral}
suggests $\hat{\upsilon}^2$ is complex for real $\khat$, which implies the
possibility of instability, regardless of the value of
$|\gamma-1|$. This is contradicts the second Solberg-Hoiland criterion 
(Eq. \ref{solberg2}), which imply instability for adiabatic
perturbations in a thin, vertically isothermal disk would require 
$|\gamma-1|\ll1$. 
%also need real freq^2 in adia limit 
This is because the VGSB retains the global radial
disk structure responsible for vertical shear
($\p_z\Omega$), but ignores it elsewhere in the linear problem. 
Nevertheless, we demonstrate below
that this inconsistency is unimportant for the VSI, which occurs 
for $\beta\ll1$.   %and small scale

\subsection{Effect of global radial gradients}
In  Fig. \ref{gcorr_compare}, we plot the effect of the global radial gradient terms
proportional to $\zeta$ in Eq. \ref{lin_mass}---\ref{lin_energy}
by calculating the
fundamental VSI growth rates using three approaches. We compute growth rates from the dispersion
relation Eq. \ref{relax_disp}, which assumes $\zeta=0$;  from
Eq. \ref{lin_mass}---\ref{lin_energy} with $\zeta=0$; and from
Eq. \ref{lin_mass}---\ref{lin_energy}  with $\zeta=1$.

@@

\begin{figure}
  \includegraphics[width=\linewidth,clip=true,trim=0cm 0.0cm 0cm
  0cm]{figures/gcorr_compare} 
  \caption{Growth rate of the fundamental VSI mode as a function of
    the thermal relaxation time $\beta$. The `Analytical (VGSB)' curve
    is calculated from the dispersion relation Eq. \ref{relax_disp};
    the `Numerical (VGSB)' curve is obtained by numerically solving
    the linear problem, Eq. \ref{lin_mass}---\ref{lin_energy}, with $\zeta=0$
    (which defines the linear VGSB equations); and the `Numerical
    (Full)' curve is obtained by numerically solving 
    Eq. \ref{lin_mass}---\ref{lin_energy}  with $\zeta=1$, which
    accounts for the radial disk structure.  
    The disk parameters are $(\gamma, \Gamma)=(1.4, 1.011)$ and
    $(p,q,h)=(-1.5,-1,0.05)$. The perturbation
    wavenumber is $\khat=30$. 
    \label{gcorr_compare}}  
\end{figure}

### 2015-04-23 application.tex

However, disturbances with large $\khat$ may be subject to viscous
decay in a real disk. The viscous timescale for a perturbation lengthscale 
$l\sim 1/k_x$ is $t_\mathrm{visc} = 1/\alpha\khat^2\Omega $, where
$\alpha$ is the standard alpha viscosity parameter
\citep{shakura73}. Setting $t_\mathrm{visc}$ to be longer than the
minimum VSI growth timescale $1/h\Omega$ implies 
$\khat^2\lesssim h/\alpha$. Inserting $h \sim 0.05$,
we find only modes with $\khat\lesssim 20$ are relevant for
$\alpha\sim 10^{-4}$ or $\khat\lesssim 70$ for $\alpha\sim
10^{-5}$. \citepalias[These low viscosity levels are required for the
VSI, see][]{nelson13}.

>>>>

However, disturbances with large $\khat$ may be subject to viscous
decay in a real disk. Adopting the standard precription for the
kinematic viscosity $\nu = \alpha c_s H$ \citep{shakura73}, where
$\alpha$ is the dimensionless viscosity parameter, 
the viscous timescale for a perturbation lengthscale 
$l\sim 1/k_x$ is $t_\mathrm{visc} = 1/\alpha\khat^2\Omega $. 
Setting $t_\mathrm{visc}$ to be longer than the
minimum VSI growth timescale $1/h\Omega$ implies 
$\khat^2\lesssim h/\alpha$. Inserting $h \sim 0.05$,
we find only modes with $\khat\lesssim 20$ are relevant for
$\alpha\sim 10^{-4}$ or $\khat\lesssim 70$ for $\alpha\sim
10^{-5}$. \citepalias[These low viscosity levels are required for the
VSI, see][]{nelson13}.

### 2015-04-23 linear_adia.tex

Eq. \ref{relax_cond_simp1}---\ref{relax_cond_simp2} is a pair of
simultaneous equations for $\hat{\omega}$ and $\beta$ for marginal
stability. (It is important to note that marginal stability is an
assumption.) If $\beta\ll 1$ and the disk is sufficiently thin so that  
$|h q \khat |^2\lesssim 1$, 
then an approximate solution to Eq. \ref{relax_cond_simp1} is 
\begin{align}
  \left(\hat{\omega}\khat\right)^2 \simeq (M+1)\left(1 -
    \frac{M|h q \khat|^2}{2M+1}\right).  \label{marg_freq_approx}
\end{align}
Note that for $q=0$, Eq. \ref{marg_freq_approx} reduces to 
the dispersion relation for low-frequency inertial waves in vertically
isothermal disks without vertical shear (see Appendix
\ref{stable_novshear}). Inserting Eq. \ref{marg_freq_approx} into
Eq. \ref{relax_cond_simp2}, we obtain 
\begin{align}\label{bcrit_gen}
  \beta &= \frac{h|q|}{(\gamma - 1)}
  \frac{\sqrt{1+M}}{(1+2M)}\left[1 + \frac{M(2M-1)|h q
      \khat|^2}{2(2M+1)^2}\right].
%  &\lesssim \beta_\mathrm{crit}\notag
\end{align}
Since $|h q \khat|^2\lesssim 1$ by assumption, we have  
\begin{align}\label{iso_vsi_cond}
  \beta\leq \beta_\mathrm{crit}  \equiv
  \frac{h|q|}{\gamma-1}, 
\end{align}
where $\beta_\mathrm{crit}$ is the critical thermal timescale for
the fundamental mode ($M=0$). 
We comment the expression for $\beta_\mathrm{crit}$ may in fact be
obtained without assuming it is much less than unity from the onset or
$|h q\khat|\lesssim 1$. % This is because for $M=0$, 
% Eq. \ref{relax_cond_simp1} --- \ref{relax_cond_simp2} simplify 
We will  
confirm $\beta_\mathrm{crit}$ numerically in \S\ref{bcrit_num_test}.

>>>>

Eq. \ref{relax_cond_simp1}---\ref{relax_cond_simp2} is a pair of
simultaneous equations for $\hat{\omega}$ and $\beta$ \emph{assuming} marginal 
stability. %  (It is important to note that marginal stability is an
% assumption.)
If $\beta\ll 1$ and the disk is sufficiently thin so that  
$|h q \khat |^2\lesssim 1$, 
then an approximate solution to Eq. \ref{relax_cond_simp1} is 
\begin{align}
  \left(\hat{\omega}\khat\right)^2 \simeq (M+1)\left(1 -
    \frac{M|h q \khat|^2}{2M+1}\right).  \label{marg_freq_approx}
\end{align}
Note that for $q=0$, Eq. \ref{marg_freq_approx} reduces to 
the dispersion relation for low-frequency inertial waves in vertically
isothermal disks without vertical shear (see Appendix
\ref{stable_novshear}). Inserting Eq. \ref{marg_freq_approx} into
Eq. \ref{relax_cond_simp2}, we obtain 
\begin{align}\label{bcrit_gen}
  \beta &= \frac{h|q|}{(\gamma - 1)}
  \frac{\sqrt{1+M}}{(1+2M)}\left[1 + \frac{M(2M-1)|h q
      \khat|^2}{2(2M+1)^2}\right].
%  &\lesssim \beta_\mathrm{crit}\notag
\end{align}
Since $|h q \khat|^2\lesssim 1$ by assumption, we have  
\begin{align}\label{iso_vsi_cond}
  \beta\leq \beta_\mathrm{crit}  \equiv
  \frac{h|q|}{\gamma-1}, 
\end{align}
where $\beta_\mathrm{crit}$ is the critical thermal timescale for
the fundamental mode ($M=0$). 
We comment the expression for $\beta_\mathrm{crit}$ may in fact be
obtained without assuming it is much less than unity from the onset or
$|h q\khat|\lesssim 1$. % This is because for $M=0$, 
% Eq. \ref{relax_cond_simp1} --- \ref{relax_cond_simp2} simplify 
We will  
confirm $\beta_\mathrm{crit}$ numerically in \S\ref{bcrit_num_test}.
